Secure Composition \\ of Robust and Optimising Compilers

### 2024-05-27 csf2/main.tex
remove inline todo
sec 3

@@

Notably, real-world compilation chains also perform a series of (sequential) passes whose main purpose is not necessarily to translate from one language to another, but to, e.g., optimise the code or enforce a certain property.
Both examples can be seen in practice, e.g. as in the work of~\cite{nagarakatte2009soft,nagarakatte2010cets,akritidis2009baggy,wegman1991ccp,manjikian1997fusion} and many more.
% \MPin{survey to cite? this is a bit informal MK: hard to find a survey on different compiler passes... I can't even find one just for compiler enforced memsafety (that looks good) or for compiler translations in general}.
% 
% Consider the compilation passes of LLVM~\cite{lattner2004llvm}, which are handled in the \texttt{PassManager} class.
% By manipulating the internal state of a \texttt{PassManager} object, the schedule of compiler optimisation passes can be hardcoded.
% That way, an order of optimisations can be found by hand that yields the best results.
% \MP{what's the point of the previous 3 sentences?(commented)}
Consider the following two LLVM optimisation passes: \gls*{cf}, which rewrites constant expressions to the constant they evaluate to, and \gls*{dce}, which removes dead code by rewriting conditional branches.
The order in which \gls*{cf} and \gls*{dce} are performed influences the final result of the compilation (see \Cref{fig:cfdceex}).
\begin{figure}[!ht]
  \begin{center}
    \begin{tikzpicture}[scale=0.62, every node/.style={transform shape}]
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left] (progunopt) {\begin{lstlisting}[language=ml,basicstyle=\small\ttfamily]
let a = true in
if a then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\large\ttfamily]
if true then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\Large\ttfamily]
print "a"
\end{lstlisting}};% the following is a hack to get my syntax highlighting to work again \end{lstlisting}
      % edges
      \draw[->,very thick] (progunopt) edge[loop left,left] node {\gls*{dce}} (progunopt);
      \draw[->,very thick] (progunopt) edge[sloped,above] node {\gls*{cf}} (progoptcf);
      \draw[->,very thick] (progoptcf) edge[sloped,above] node {\gls*{dce}} (progoptcfdce);
    \end{tikzpicture}
  \end{center}
  \caption{Example program where the level of optimisations differ for one pass of applying \gls*{cf} and \gls*{dce} in any order. %
  Every edge is a compilation pass and the label on the edge states what the pass does, i.e., \gls*{cf} or \gls*{dce}. %
  The source code in the nodes is a glorified compiler intermediate representation and the code gets more optimised towards the right hand side of the figure.}\label{fig:cfdceex}
\end{figure}
This {\em phase ordering problem} is well--known in literature and a practical solution is to simply perform a fixpoint iteration of the optimisation pipeline~\cite{click1995combining}.
Hence, it is of interest to analyze the security of compilation passes and their interaction within a compilation pipeline.

For starters, suppose there are traces $\src{\trace}$ and $\trg{\trace}$ as well as two trace relations $\sim_1$ and $\sim_2$. 
The traces are related $\src{\trace}\sim_1\bullet\sim_2\trg{\trace}$ if there is another trace $\irl{\trace}$ such that $\src{\trace}\sim_1\irl{\trace}$ and $\irl{\trace}\sim_2\trg{\trace}$.

\bul{A relation $\sim$ is well-formed w.r.t. a class of target-level properties $\trg{\class}$} iff \rul{the universal image preserves set membership}.
\begin{definition}[Well-formedness of $\sim$ for a Class $\trg{\class}$]\label{def:wfc:sig:tracerel}
  % $$

@@

With these notions of well-formedness, it is now possible to prove the main result that secure compilers in the robust compilation framework~\cite{abate2021extacc} compose {\em sequentially}.

@@

Since the composition of secure compilers is again a secure compiler, the theorems generalise to a whole chain of $n$ secure compilers. 
Moreover, the previously defined theory dually holds for the existential image $\tau_\sim\left(\src{\pi}\right)$, but these results have been moved to the technical report in the interest of space.

@@

\Cref{corr:rtp} provides an easy to use way to compose secure compilers without any caveats. 
However, while \Cref{thm:rtpsim:sig} explicitly requires that $\sim_1$ is well-formed w.r.t. $\trg{\class[_2]}$, if $\sim_2$ is not well-formed w.r.t. $\trg{\class[_1]}$, care must be taken. 
%Worst-case is that the resulting class $\sigma_{\sim_1\bullet\sim_2}\left(\trg{\class[_1]}\cap\trg{\class[_2]}\right)$ becomes empty, even though $\trg{\class[_1]}\cap\trg{\class[_2]}$ is not necessarily empty.
\Cref{subsec:compatsecpasses} provides an example of this phenomenon.

@@

\subsection{Secure Upper and Lower Composition}\label{sec:other-compos}
Besides sequential composition, there are two other compositions, namely an {\em upper}, i.e., a compiler that takes multiple inputs and yields one output, and a {\em lower} composition, i.e., a compiler that takes one input and yields multiple outputs.
{Define the upper composition $\cc{\src{L}+\obj{L}}{\trg{L}}$} as follows:
Given a program \texttt{p}, its compiled counterpart is obtained by {plugging \texttt{p} into $\cc{\src{L}}{\trg{L}}$ if $\texttt{p}\in\src{L}$} or by {plugging \texttt{p} into $\cc{\obj{L}}{\trg{L}}$ if $\texttt{p}\in\obj{L}$}.
\begin{definition}[Upper Composition]
  $$\text{{$\cc{\src{L}+\obj{L}}{\trg{L}}$}}\isdef
  % Given \texttt{p}, yield
  \lambda \texttt{p}\ldotp
  \left\{\mbox{\begin{tabular}{c}
    {{if $\texttt{p}\in\src{L}$, then $\cc{\src{L}}{\trg{L}}(\texttt{p})$}} \\
    \mbox{{if $\texttt{p}\in\obj{L}$, then $\cc{\obj{L}}{\trg{L}}(\texttt{p})$}} \\
  \end{tabular}}\right.$$
%
%
%  $\begin{cases}
%                                                                            \rul{\cc{\src{L}}{\trg{L}}(\texttt{p}) & \text{if }\texttt{p}\in\src{L}} \\
%                                                                            \cc{\obj{L}}{\trg{L}}(\texttt{p}) & \text{if }\texttt{p}\in\obj{L}
%                                                                          \end{cases}$
\end{definition}

@@

Future work intends to use upper and lower composition as means to build a secure compiler incrementally.

@@

This section introduces trace properties of interest for this paper: \gls*{tms}, \gls*{sms}, \gls*{ms}, \gls*{scct}, and \gls*{ss}.
These properties are of practical importance (as mentioned in \Cref{sec:introduction}) and also of interest in the case study (\Cref{sec:casestud:defs,sec:casestud:rtp}) presented later. 
While up to \gls*{scct}, none of the properties is novel, for self-containment of the paper, their definition is necessary for the case study.
For each of the presented properties, the technical appendix accompanying the paper defines corresponding monitors, which refine a given property and have a key role in the proofs of this paper.

@@

Note that \Cref{def:trace:msdef} ignores data isolation, there may still be memory-safety issues introduced by side-channels.

\subsection{Memory Safety Trace Model with Constant Timeness}\label{subsec:scct:tracemodel}

This section starts off by extending the memory safety trace model (\Cref{subsec:basic:memsafety:tracemodel}) with a {\em security tag} ($\securitytag$) that indicates whether the involved event contains sensitive information ($\lock$) or not ($\unlock$).

\vspace{-1em}
\[
  \begin{array}{rrcl}
    (\text{Base-Events}) & \preevent[_{\ctsafe}] &:=& \preevent[_{\mssafe}] \mid \ev{Branch\ }n \mid \ev{Binop\ }n\\
    (\text{Security Tags}) & \securitytag &:=& \lock \mid \unlock\\ 
    (\text{Events}) & \event[_{\ctsafe}] &:=& \preevent[_{\ctsafe}];\securitytag \mid \emptyevent \mid \lightning \\ 
  \end{array}
\]

For cryptographic code, there is a general guideline that secrets must not be visible on a trace~\cite{ctguidelines}, i.e., marked as $\unlock$.
In turn, an instruction whose timing is data-dependent must not have a secret as an operand.
Typical operations with data-dependent timing are branches and certain binary operations, such as division. 
\footnote{
	This is architecture-dependent, but division is an operation that serves as a classic example for a data-dependent timing instruction~\cite[p.~755]{arm-refman}.
}
Both operations are also modelled in the trace model by extending the set of base-events with branches ($\ev{Branch\ n}$) and binary operations ($\ev{Binop\ n}$).

>>>>

\smallskip%MP: the underline compacts the sections too much

@@

Notably, real-world compilation chains also perform a series of (sequential) passes whose main purpose is not necessarily to translate from one language to another, but to, e.g., optimise the code or enforce a certain property.
Both examples can be seen in practice, e.g.,~\cite{nagarakatte2009soft,nagarakatte2010cets,akritidis2009baggy,wegman1991ccp,manjikian1997fusion} and many more.
% \MPin{survey to cite? this is a bit informal MK: hard to find a survey on different compiler passes... I can't even find one just for compiler enforced memsafety (that looks good) or for compiler translations in general}.
% 
% Consider the compilation passes of LLVM~\cite{lattner2004llvm}, which are handled in the \texttt{PassManager} class.
% By manipulating the internal state of a \texttt{PassManager} object, the schedule of compiler optimisation passes can be hardcoded.
% That way, an order of optimisations can be found by hand that yields the best results.
% \MP{what's the point of the previous 3 sentences?(commented)}
Consider the following two LLVM optimisation passes: \gls*{cf}, which rewrites constant expressions to the constant they evaluate to, and \gls*{dce}, which removes dead code by rewriting conditional branches.
The order in which \gls*{cf} and \gls*{dce} are performed influences the final result of the compilation (see \Cref{fig:cfdceex}).
\begin{figure}[!ht]
  \begin{center}
    \begin{tikzpicture}[scale=0.62, every node/.style={transform shape}]
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left] (progunopt) {\begin{lstlisting}[language=ml,basicstyle=\small\ttfamily]
let a = true in
if a then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progunopt] (progoptcf) {\begin{lstlisting}[language=ml,basicstyle=\large\ttfamily]
if true then
  print "a"
else
  print "b"
\end{lstlisting}};
      \node[minimum width=3cm,minimum height=2.5cm,draw=black,very thick,align=left,right=1.0 of progoptcf] (progoptcfdce) {\begin{lstlisting}[language=ml,basicstyle=\Large\ttfamily]
print "a"
\end{lstlisting}};% the following is a hack to get my syntax highlighting to work again \end{lstlisting}
      % edges
      \draw[->,very thick] (progunopt) edge[loop left,left] node {\gls*{dce}} (progunopt);
      \draw[->,very thick] (progunopt) edge[sloped,above] node {\gls*{cf}} (progoptcf);
      \draw[->,very thick] (progoptcf) edge[sloped,above] node {\gls*{dce}} (progoptcfdce);
    \end{tikzpicture}
  \end{center}
  \caption{Example program where the level of optimisations differ for one pass of applying \gls*{cf} and \gls*{dce} in any order. %
  Every edge is a compilation pass and the label on the edge states what the pass does, i.e., \gls*{cf} or \gls*{dce}. %
  The source code in the nodes is a glorified compiler intermediate representation and the code gets more optimised towards the right hand side of the figure.}\label{fig:cfdceex}
\end{figure}
This {\em phase ordering problem} is well--known in literature and a practical solution is to simply perform a fixpoint iteration of the optimisation pipeline~\cite{click1995combining}.
Hence, it is of interest to analyze the security of compilation passes and their interaction within a compilation pipeline.

\smallskip

The key notions for our main results are the definition of a trace relation, and the definition of when is a trace relation well-formed with respect to a class.
%
Consider traces $\src{\trace}$ and $\trg{\trace}$ as well as two trace relations $\sim_1$ and $\sim_2$. 
% 
The traces are related $\src{\trace}\sim_1\bullet\sim_2\trg{\trace}$ if there is another trace $\irl{\trace}$ such that $\src{\trace}\sim_1\irl{\trace}$ and $\irl{\trace}\sim_2\trg{\trace}$.
% 
\bul{A relation $\sim$ is well-formed w.r.t. a class of target-level properties $\trg{\class}$} iff \rul{the universal image preserves set membership}.
\begin{definition}[Well-formedness of $\sim$ for a Class $\trg{\class}$]\label{def:wfc:sig:tracerel}
  % $$

@@

We can now state our main result: secure compilers in the robust compilation framework~\cite{abate2021extacc} compose {\em sequentially}. 
%

@@

Since the composition of secure compilers is again a secure compiler, the theorems generalise to a whole chain of $n$ secure compilers. 
% 
Note that the previous results also hold for the existential image $\tau_\sim\left(\src{\pi}\right)$, but these results have been moved to the technical report in the interest of space.

@@

\Cref{corr:rtp} provides an easy way to compose secure compilers without well-formedness of trace relations. 
% 
However, while \Cref{thm:rtpsim:sig} explicitly requires that $\sim_1$ is well-formed w.r.t. $\trg{\class[_2]}$, if $\sim_2$ is not well-formed w.r.t. $\trg{\class[_1]}$, care must be taken. 
%Worst-case is that the resulting class $\sigma_{\sim_1\bullet\sim_2}\left(\trg{\class[_1]}\cap\trg{\class[_2]}\right)$ becomes empty, even though $\trg{\class[_1]}\cap\trg{\class[_2]}$ is not necessarily empty.
This is further discussed in \Cref{subsec:compatsecpasses}.

@@

\subsection{Secure Upper and Lower Composition}\label{sec:other-compos}
Besides sequential composition, there are two other compositions, namely an {\em upper}, i.e., a compiler that takes multiple inputs and yields one output, and a {\em lower} composition, i.e., a compiler that takes one input and yields multiple outputs.
We {fefine the upper composition $\cc{\src{L}+\obj{L}}{\trg{L}}$} as follows:
Given a program \texttt{p}, its compiled counterpart is obtained by {plugging \texttt{p} into $\cc{\src{L}}{\trg{L}}$ if $\texttt{p}\in\src{L}$} or by {plugging \texttt{p} into $\cc{\obj{L}}{\trg{L}}$ if $\texttt{p}\in\obj{L}$}.
\begin{definition}[Upper Composition]
  $$\text{{$\cc{\src{L}+\obj{L}}{\trg{L}}$}}\isdef
  % Given \texttt{p}, yield
  \lambda \texttt{p}\ldotp
  \left\{\mbox{\begin{tabular}{c}
    {{if $\texttt{p}\in\src{L}$, then $\cc{\src{L}}{\trg{L}}(\texttt{p})$}} \\
    \mbox{{if $\texttt{p}\in\obj{L}$, then $\cc{\obj{L}}{\trg{L}}(\texttt{p})$}} \\
  \end{tabular}}\right.$$
%
%
%  $\begin{cases}
%                                                                            \rul{\cc{\src{L}}{\trg{L}}(\texttt{p}) & \text{if }\texttt{p}\in\src{L}} \\
%                                                                            \cc{\obj{L}}{\trg{L}}(\texttt{p}) & \text{if }\texttt{p}\in\obj{L}
%                                                                          \end{cases}$
\end{definition}

@@

Future work intends to use upper and lower composition as means to build a secure compiler incrementally.
\MPin{
	we should add briefly which real-world compilers (taken from the elided examples) are upper and lower compos
}
\MP{no coq here ?}

@@

This section introduces trace properties of interest for this paper: \gls*{tms}, \gls*{sms}, \gls*{ms}, \gls*{scct}, and \gls*{ss}.
These properties are of practical importance (as mentioned in \Cref{sec:introduction}) and also of interest in the case study presented later (\Cref{sec:casestud:defs,sec:casestud:rtp}). 
While up to \gls*{scct}, none of the properties is novel, for self-containment of the paper, their definition is necessary for the case study.
For each of the presented properties, the technical appendix accompanying the paper defines corresponding monitors, which refine a given property and have a key role in the proofs of this paper.

@@

Note that \Cref{def:trace:msdef} ignores data isolation, so there may still be memory-safety issues introduced by side-channels.

\subsection{Memory Safety Trace Model with Constant Time}\label{subsec:scct:tracemodel}

This section starts by extending the memory safety trace model (\Cref{subsec:basic:memsafety:tracemodel}) with a {\em security tag} ($\securitytag$) that indicates whether the involved event contains sensitive information ($\lock$) or not ($\unlock$).

\vspace{-.5em}
\[
  \begin{array}{rrcl}
    (\text{Base-Events}) & \preevent[_{\ctsafe}] &:=& \preevent[_{\mssafe}] \mid \ev{Branch\ }n \mid \ev{Binop\ }n\\
    (\text{Security Tags}) & \securitytag &:=& \lock \mid \unlock\\ 
    (\text{Events}) & \event[_{\ctsafe}] &:=& \preevent[_{\ctsafe}];\securitytag \mid \emptyevent \mid \lightning \\ 
  \end{array}
\]

For cryptographic code, there is a general guideline that secrets must not be visible on a trace~\cite{ctguidelines}, i.e., marked as $\unlock$.
In turn, an instruction whose timing is data-dependent must not have a secret as an operand.
Typical operations with data-dependent timing are branches and certain binary operations, such as division.%
\footnote{
	This is architecture-dependent, but division is an operation that serves as a classic example for a data-dependent timing instruction~\cite[p.~755]{arm-refman}.
}
Both operations are represented in the trace model by extending the set of base-events with branches ($\ev{Branch\ n}$) and binary operations ($\ev{Binop\ n}$).
